Bayesian multi-proxy reconstruction of early Eocene latitudinal temperature gradients

### 2023-09-22 manuscript/main.qmd
reviewer 2 minor notes corrected

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012], using a nearest-living-relative approach [e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1\; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained. For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N--32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, placing 97.5% of the probability density above the minimum. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical, lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene. Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures being between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is ascribed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median and 95% CI of temperatures where then taken from all temperature estimates obtained at the latitudes of interest.

The latitudinal gradient is calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude). To facilitate comparison with earlier estimates, we also calculate the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles. Given the sigmoidal shape of the modern as well as the Eocene gradient (see Fig. 4), these results are broadly comparable to a gradient inferred from the zonal average of equatorial and high-latitude temperatures, as has been done in some earlier studies [@Evans2018].

>>>>

We apply this model to the record of the early Eocene climatic optimum (EECO), combining a compilation of geochemical proxies [@Hollis2019], mangrove communities [@Popescu2021], and coral reefs [@Zamagni2012]. We use a nearest-living-relative approach [e.g. @Greenwood2017] to establish broad temperature ranges for the ecological data. We choose the EECO to demonstrate the application of the model due to its significance as the interval with the warmest sustained temperatures of the Cenozoic [@Pross2012], rendering it a potential analogue for extreme climate warming scenarios [@Burke2018]. Our integrative approach allows us to shed new light on the long-standing dispute on the steepness of the early Eocene temperature gradient [Table 1\; @Sloan1990; @Markwick1994; @Huber2011; @Tierney2017; @Inglis2020].

@@

Geochemical climate proxy data were extracted from a latest Paleocene and early Eocene compilation [@Hollis2019]. This compilation provides sea surface temperature data on four different geochemical proxies for reconstructing seawater temperature: &delta;^18^O, &Delta;~47~, Mg/Ca and TEX~86~. For our analyses, this dataset was restricted to the EECO (defined as 53.8--49.1 Ma) and samples originating from near the ocean surface or mixed layer. Samples labelled as "thermocline" or "sub-thermocline" were excluded. Recrystallised &delta;^18^O samples were also excluded as secondary diagenetic calcite precipitated after deposition can bias isotope measurements and offset temperature values [@Schrag1999]. This filtering resulted in most &delta;^18^O samples being excluded from the dataset (retaining 8 out of 152). After data filtering, `r n_geochem` geochemical proxy samples from `r n_geochem_location` locations remained (Fig. 1). For a detailed description of each proxy see @Hollis2019.

@@

**Coral reefs.** Today, shallow warm-water coral reefs are limited to tropical and subtropical latitudes (\~34&deg;&nbsp;N--32&deg;&nbsp;S), with minimum sea surface temperature tolerances (\~18&deg;C) being the primary constraint on this distribution [@johannes1983; @kleypas1999; @yamano2001]. As coral reefs reside at the upper thermal limit of the oceans today, their maximum sea surface temperature tolerance is less well-constrained, with some studies suggesting up to 35.6&deg;C in the geological past [@jones2022reef]. Nevertheless, coral reefs have frequently been recognised as tracers of past (sub-)tropical conditions [@ziegler1984; @kiessling2001]. During the Eocene, coral communities and reefs expanded across tropical and temperate latitudes, with communities found up to palaeolatitudes of 43&deg;&nbsp;N [@Zamagni2012]. Using a compilation of Paleocene--early Eocene coral reefs and community localities [@Zamagni2012], we generated quantitative sea surface temperature estimates for the ECCO. To do so, we extracted localities from the compilation that are inferred to be Ilerdian (early Eocene) coral reefs, and that could be confidently assigned to the EECO. We excluded coral knobs and coral-bearing mounds which might have broader climatic limits than warm-water coral reef ecosystems. This filtering resulted in four unique coral reef localities remaining for the EECO, all of which conform to the modern latitudinal range of coral reefs (\<34&deg;&nbsp;N). Subsequently, we used statistically derived temperature limits (minimum = 21&deg;C, average = 27.6&deg;C, maximum = 29.5&deg;C) from the published literature [@kleypas1999] to define a normal probability distribution of potential temperature values for coral reef localities. This normal probability distribution was defined with a mean of 27.6&deg;C and a standard deviation of 2.125&deg;C, which places the minimum (21&deg;C) at the lower end of the 95% highest density interval of that distribution. As the distribution of modern corals is skewed towards warmer temperatures, this approach results in 16.5% of the probability being placed on temperatures \> 29.5&deg;C, allowing for the possibility that Eocene coral reefs were adapted to warmer conditions than present-day coral reefs.

**Mangroves.** Mangroves are distributed throughout the tropics and subtropics today. While factors besides sea surface temperatures (SST) influence the distribution of mangroves, empirical lower temperature limits have been established for the genera *Avicennia* (15.6&deg;C) and *Rhizophora* (20.7&deg;C) [@Quisthoudt2012]. Both *Avicennia* and members of the Rhizophoraceae family were widespread and co-occurred across tropical and temperate latitudes in the early Eocene. Only *Avicennia*, however, occurred at polar latitudes [@Suan2017; @Popescu2021]. Assuming that Eocene members of these mangrove taxa conform to similar climatic requirements as their modern relatives, the presence and absence of *Avicennia* and Rhizophoraceae pollen can be used as a palaeotemperature indicator. For this analysis, published mangrove occurrence data were taken from @Popescu2021, and converted to quantitative temperature estimates. From this data, we identify two types of pollen assemblages which we ascribe different temperature distributions:

1)  *Avicennia*-only assemblages ($n=2$): the absence of Rhizophoraceae is indicative of temperatures between 15.6&deg;C (lower temperature limt of *Avicennia*) and 20.7&deg;C (lower temperature limit of *Rhizophora*). However, a value of 22.5&deg;C is assumed as the upper temperature limit here as *Rhizophora* is rare below this temperature. We define the *Avicennia*-only temperature distribution as a normal distribution with a mean of 19.05&deg;C and a standard deviation of 1.725&deg;C, resulting in 95% of the probability density being placed within the temperature limits.

@@

```{=html}
<!--
- Calculate lat gradient
- Calculate global T
[Fig. 4]
-->
```
Modelled sea surface temperature estimates were generated with [@eq-nu], calculating the sea surface temperatures at any latitude with the parameter estimates of each iteration from the posterior. The median and 95% CI of temperatures were then taken from all temperature estimates obtained at the latitudes of interest.

The latitudinal gradient was calculated as the difference between the modelled temperature at the equator (0&deg; latitude) and at the poles (90&deg; absolute latitude). To facilitate comparison with earlier estimates, we also calculated the gradient with the temperature at the polar circle (66.6&deg; absolute latitude) being used instead of the temperature at the poles.
